---
title: "R & RStudio: Installation and Updates"
author: "DAPR Team @ UoE Psychology"
toc: true
format:
  html:
    theme: cosmo
editor: visual
---

## Introduction

R is the name of both:

-   the R programming language, i.e. the language you use to tell the PC what computations to perform; and
-   the R interpreter, which actually executes the computations written in the R language.

RStudio is a user-friendly interface around R that makes the programming experience much nicer. It bundles together

-   a file editor panel,
-   a console panel with the R interpreter,
-   an environment panel showing all the objects you have created
-   a files and plots panel to see your PC files and the plots you create.

The programming language R comes with built-in functionality, like computing the square root with the `sqrt()` function. Over the years, R users have created lots of additional functionalities that is available on demand. If you particular application requires extra functionality not available with the basic R installation, you can add it by *installing packages*. A package is an additional bit of functionality and it needs to be installed only once to bring it from the web to your own R installation. However, it is also important to keep packages updated once or twice a year.

This document explains how to

A)  install R and RStudio, if you never used them before;
B)  update R and RStudio, if you have old versions;
C)  how to update your R packages.

## A. Installing R and RStudio {#install-rstudio}

::::: panel-tabset
### Windows PC

1.  If you have a previous R installation, please uninstall the following software from your PC if you have it:
    -   R
    -   Rtools
    -   RStudio
2.  Install R
    -   Go to <https://cran.r-project.org/bin/windows/base/>
    -   Click **Download R-4.5.1 for Windows**
    -   Open the installer and follow the instructions while keeping the default options
3.  Install RTools
    -   Go to <https://cran.r-project.org/bin/windows/Rtools/>
    -   Click **RTools 4.5**
    -   Click **Rtools45 installer**
    -   Open the installer and follow the instructions while keeping the default options
4.  Install RStudio
    -   Go to <https://posit.co/download/rstudio-desktop/>, ignore the "install R" stuff (you've done that already). In the table of RStudio installers, find the one for your operating system and download it.\
    -   Open the installer and follow the instructions while keeping the default options.

### Apple macOS PC

1.  If you have a previous R installation, please uninstall the following software from your PC if you have it:
    -   R
    -   XQuartz
    -   RStudio
2.  Install R
    -   Check which Apple PC you have. On the top-left of your screen click the Apple logo, click About This Mac, note what the Processor line says.
    -   Go to <https://cran.r-project.org/bin/macosx/> and download:
        -   `R-4.5.1-arm64.pkg` if you have an Apple silicon arm64 macOS PC.
        -   `R-4.5.1-x86_64.pkg` if you have an Intel-based macOS PC
    -   Open the installer and follow the instructions while keeping the default options.
3.  Install XQuartz
    -   Download XQuartz: <https://www.xquartz.org/>
    -   Open the installer and follow the instructions while keeping the default options
4.  Install RStudio
    -   Go to <https://posit.co/download/rstudio-desktop/>, ignore the "install R" stuff (you've done that already). In the table of RStudio installers, find the one for your operating system and download it.\
    -   Open the installer and follow the instructions while keeping the default options.

### Chromebooks/Ipads etc

You cannot install R or RStudio on Chromebooks or Ipads.

::: {.callout-important icon="false"}
#### During week 1 of the course

In your first week of DAPR2/3/USMR, you should access the RStudio Server provided by Noteable. To access it follow the steps below:

a.  Go to the course LEARN page
b.  Scroll to the bottom of the LEARN page
c.  Click "RStudio for Chromebook/Ipad Users (Noteable)"
d.  Select "RStudio" from the dropdown menu
e.  Click "Start"
:::

::: {.callout-note collapse="true" icon="false"}
#### From week 2 onwards

You can use RStudio on the PPLS Server.

**First try these steps:**

1.  Log in to [EASE](https://www.ease.ed.ac.uk/) using your university UUN and password

2.  Set your RStudio password [here](https://rstudio.ppls.ed.ac.uk/setpass.php), the username will be the same as your UUN (make sure you type your UUN correctly!)

3.  Access the server from <https://rstudio.ppls.ed.ac.uk> using your university UUN and the password you set above in (2).

**If that didn't work:**

Please fill out [this form](https://forms.office.com/r/Pc5aqhyveE) and wait for an email from us with instructions.
:::
:::::

## B. Updating R and RStudio

### B.1 Updating R

If your R version or packages are old, you might get errors or find that the code required to run some examples is different from the one provided.

Recall that R is the underlying programming language, while RStudio is just a wrapper around R providing a nicer user interface (text editor, file explorer and graphical window). Updating RStudio (the interface) will not update the R programming language which runs your code.

#### Check which R version you have

Check that your R version is up-to-date by typing into the R console:

```         
version
```

After pressing Enter, you should see something like this:

```         
platform       << this will be different for each of you >>     
arch           << this will be different for each of you >>                  
os             << this will be different for each of you >>     
system         << this will be different for each of you >>        
status                                     
major          4                           
minor          5.1                         
<< and more lines which you can ignore... >>
```

Please check here: https://www.r-project.org/ to see what the most recent version is. If you have a different number for "major" than the most recent, it may well be time for an update.

#### Updating R

::: panel-tabset
##### Windows PC

Open RStudio, and type the following in the R console. Press Enter to run the code.

```{r eval=FALSE}
install.packages("installr")
installr::updateR()
```

##### Apple macOS PC

1.  Follow step 2 from the instructions in [section A](#install-r) in order to install the newest version of R.\
2.  In RStudio, look in the top menu for Tools \> Global Options. In the "R General" tab, you can choose which option of R you want to use. Make sure this is set to the one you just downloaded.
:::

### B.2 Updating RStudio

Check if you need to update RStudio by clicking **Help** and then **Check for updates**.

If an update is available, follow step 4 in [section A](#install-rstudio).

## C. Updating R packages {#update-packages}

Open RStudio. In the R console, type the following and press Enter:

```{r eval=FALSE}
options(pkgType = "binary")
update.packages(ask = FALSE)
```

## History

-   Last updated by UN in September 2025 for R version 4.5.1
-   Last updated by UN in May 2024 for R version 4.4.0
-   Last updated by UN in September 2023 for R version 4.3.1
-   Last updated by UN in November 2022 for R version 4.2.2
-   Last updated by UN in July 2022 for R version 4.2.1

Department of Psychology\
The University of Edinburgh
