---
title: "Updating R & Packages"
author: "uoepsy stats team"
output:
  html_document:
    theme: readable
    toc: true
    toc_depth: 1
---


R is the name of both:

- the R programming language, i.e. the language you use to tell the PC what computations to perform; and
- the R interpreter, which actually executes the computations written in the R language.

RStudio is a user-friendly interface around R that makes the programming experience much nicer. It bundles together 

- a file editor panel, 
- a console panel with the R interpreter, 
- an environment panel showing all the objects you have created
- a files and plots panel to see your PC files and the plots you create.


The programming language R comes with built-in functionality, like computing the square root with the `sqrt()` function. Over the years, R users have created lots of additional functionalities that is available on demand. If you particular application requires extra functionality not available with the basic R installation, you can add it by *installing packages*. A package is an additional bit of functionality and it needs to be installed only once to bring it from the web to your own R installation. However, it is also important to keep packages updated once or twice a year.


This document explains how to 

a) install R and RStudio, if you never used them before; 
b) update R and RStudio, if you have old versions;
c) how to update your R packages.


---


# A. Installing R and RStudio

## A.1 Installing R {.tabset .tabset-fade .tabset-pills}

### Windows PC

#### Uninstall any previous installations

If you have a previous R installation, please uninstall the following software from your PC if you have it:

- R
- Rtools

#### Install R

Go to <https://cran.r-project.org/bin/windows/base/> and click `Download R << number >> for Windows`.

Open the installer and follow the instructions.

#### Install RTools

Go to <https://cran.r-project.org/bin/windows/Rtools/> and under the section `Installing Rtools`, download the version that matches your PC: either Windows 64-bit (most likely) or Windows 32-bit (if your PC is very old).

Open the installer and follow the instructions.



### Apple macOS PC

#### Uninstall any previous installations

If you have a previous R installation, please go to the Applications folder in your Finder window and move the following applications to the Bin:

- R
- XQuartz


#### Install R

Check which Apple PC you have. On the top-left of your screen click the Apple logo, click About This Mac, note what the Processor line says. If it has the word "Intel", you want to follow 1 below, and 2 otherwise.

Go to <https://cran.r-project.org/bin/macosx/> and download one of these: 

1. `R-4.1.1.pkg (notarized and signed)` if you have an Intel-based macOS PC
2. `R-4.1.1-arm64.pkg (notarized and signed)` if you have an Apple silicon arm64 macOS PC.

Open the installer and follow the instructions.


#### Install XQuartz

Download XQuartz: [https://xquartz.macosforge.org/](https://xquartz.macosforge.org/). 

Open the installer and follow the instructions.



### Chromebooks

You cannot install R on Chromebooks. Go directly to Installing RStudio to follow the instructions.



## A.2 Installing RStudio {.tabset .tabset-fade .tabset-pills}

### Windows PC {#install-win}

Go to <https://www.rstudio.com/products/rstudio/download/#download> and click the big blue download button.

Open the installer and follow the instructions.

### Apple macOS PC {#install-mac}

Go to <https://www.rstudio.com/products/rstudio/download/#download> and click the big blue download button.

Open the installer and follow the instructions.


### Chromebooks {#install-chrome}

1. Log in to EASE using your university UUN and password

2. Set your RStudio password [here](https://rstudio.ppls.ed.ac.uk/setpass.php), the username will be the same as your UUN (make sure you type your UUN correctly!)

3. If you believe that your username has not been granted access, please fill out the form [here](https://forms.office.com/r/Pc5aqhyveE) and we will get back to you as soon as we can.

Go to <https://rstudio.ppls.ed.ac.uk> to log in to the server using your UUN and the password you just set.


## {-}


---


# B. Updating R and RStudio

If your R version or packages are old, you might get errors or find that the code required to run some examples is different from the one provided.

Recall that R is the underlying programming language, while RStudio is just a wrapper around R providing a nicer user interface (text editor, file explorer and graphical window). Updating RStudio (the interface) will not update the R programming language which runs your code.  

## B.1 Check which R version you have

Check that your R version is up-to-date by typing into the R console:

```
version
```

After pressing Enter, you should see  something like this:

```
platform       << this will be different for each of you >>     
arch           << this will be different for each of you >>                  
os             << this will be different for each of you >>     
system         << this will be different for each of you >>        
status                                     
major          4                           
minor          1.1                         
<< and more lines which you can ignore... >>
```

Please check here: https://www.r-project.org/ to see what the most recent version is. If you have a different number for "major" than the most recent, it may well be time for an update.  


## 2. Update R {.tabset .tabset-fade .tabset-pills}

### Windows PC

Open RStudio, and type the following in the R console. Press Enter to run the code.

```{r eval=FALSE}
install.packages("installr")
installr::updateR()
```


### Apple macOS PC

To update R, you need to remove any old R installations and reinstall R. Follow the instructions from [A.1](#install-mac).


### Chromebooks

You cannot install or update R on your Chromebook. You need to access R via the RStudio Server provided by our school, and the updates are managed by the school, you don't need to do anything.

You still may want to update packages, though. Check [section C](#update-pkgs) for that.

## {-}


---


# C. Updating R packages {#update-pkgs}

Open RStudio. In the R console, type the following and press Enter:

```{r eval=FALSE}
options(pkgType = "binary")
update.packages(ask = FALSE)
```

<br>

