---
title: "R & RStudio"
subtitle: "Installation and updates"
author: "UoE Psychology"
output:
  rmdformats::material
    # theme: cosmo
    # toc: true
    # toc_depth: 2
---


# Introduction

R is the name of both:

- the R programming language, i.e. the language you use to tell the PC what computations to perform; and
- the R interpreter, which actually executes the computations written in the R language.

RStudio is a user-friendly interface around R that makes the programming experience much nicer. It bundles together 

- a file editor panel, 
- a console panel with the R interpreter, 
- an environment panel showing all the objects you have created
- a files and plots panel to see your PC files and the plots you create.


The programming language R comes with built-in functionality, like computing the square root with the `sqrt()` function. Over the years, R users have created lots of additional functionalities that is available on demand. If you particular application requires extra functionality not available with the basic R installation, you can add it by *installing packages*. A package is an additional bit of functionality and it needs to be installed only once to bring it from the web to your own R installation. However, it is also important to keep packages updated once or twice a year.


This document explains how to 

A) install R and RStudio, if you never used them before; 
B) update R and RStudio, if you have old versions;
C) how to update your R packages.





# A. Installing R and RStudio {.tabset .tabset-fade .tabset-pills #install-r}

## Windows PC

1. If you have a previous R installation, please uninstall the following software from your PC if you have it:
    - R
    - Rtools
    - RStudio
2. Install R
    - Go to <https://cran.r-project.org/bin/windows/base/> 
    - Click __Download R-4.5.0 for Windows__
    - Open the installer and follow the instructions while keeping the default options
3. Install RTools
    - Go to <https://cran.r-project.org/bin/windows/Rtools/> 
    - Click __RTools 4.5__
    - Click __64-bit ARM Rtools45 installer__
    - Open the installer and follow the instructions while keeping the default options
4. Install RStudio
    - Go to <https://posit.co/download/rstudio-desktop/>, ignore the "install R" stuff (you've done that already). In the table of RStudio installers, find the one for your operating system and download it.  
    - Open the installer and follow the instructions while keeping the default options.


## Apple macOS PC

1. If you have a previous R installation, please uninstall the following software from your PC if you have it:
    - R
    - XQuartz
    - RStudio
2. Install R
    - Check which Apple PC you have. On the top-left of your screen click the Apple logo, click About This Mac, note what the Processor line says. 
    - Go to <https://cran.r-project.org/bin/macosx/> and download:  
        - `R-4.5.0-x86_64.pkg` if you have an Intel-based macOS PC
        - `R-4.5.0-arm64.pkg` if you have an Apple silicon arm64 macOS PC.
    - Open the installer and follow the instructions while keeping the default options.
3. Install XQuartz
    - Download XQuartz: <https://www.xquartz.org/>
    - Open the installer and follow the instructions while keeping the default options
4. Install RStudio
    - Go to <https://posit.co/download/rstudio-desktop/>, ignore the "install R" stuff (you've done that already). In the table of RStudio installers, find the one for your operating system and download it.  
    - Open the installer and follow the instructions while keeping the default options.


## Chromebooks/Ipads etc

You cannot install R or RStudio on Chromebooks or Ipads.   
However, you can use RStudio on the PPLS Server.  

__First try these steps:__

1. Log in to [EASE](https://www.ease.ed.ac.uk/) using your university UUN and password

2. Set your RStudio password [here](https://rstudio.ppls.ed.ac.uk/setpass.php), the username will be the same as your UUN (make sure you type your UUN correctly!)

3. Access the server from <https://rstudio.ppls.ed.ac.uk> using your university UUN and the password you set above in (2).

__If that didn't work:__

Please fill out [this form](https://forms.office.com/r/Pc5aqhyveE) and wait for an email from us with instructions.






# B. Updating R and RStudio

## B.1 Updating R

If your R version or packages are old, you might get errors or find that the code required to run some examples is different from the one provided.

Recall that R is the underlying programming language, while RStudio is just a wrapper around R providing a nicer user interface (text editor, file explorer and graphical window). Updating RStudio (the interface) will not update the R programming language which runs your code.  

### Check which R version you have

Check that your R version is up-to-date by typing into the R console:

```
version
```

After pressing Enter, you should see  something like this:

```
platform       << this will be different for each of you >>     
arch           << this will be different for each of you >>                  
os             << this will be different for each of you >>     
system         << this will be different for each of you >>        
status                                     
major          4                           
minor          5.0                         
<< and more lines which you can ignore... >>
```

Please check here: https://www.r-project.org/ to see what the most recent version is. If you have a different number for "major" than the most recent, it may well be time for an update.  


### Updating R {.tabset .tabset-fade .tabset-pills}

#### Windows PC

Open RStudio, and type the following in the R console. Press Enter to run the code.

```{r eval=FALSE}
install.packages("installr")
installr::updateR()
```


#### Apple macOS PC

1. Follow step 2 from the instructions in [section A](#install-r) in order to install the newest version of R.  
2. In RStudio, look in the top menu for Tools > Global Options. In the "R General" tab, you can choose which option of R you want to use. Make sure this is set to the one you just downloaded.  



## B.2 Updating RStudio

Check if you need to update RStudio by clicking __Help__ and then __Check for updates__.

If an update is available, follow step 4 in [section A](#install-rstudio).



# C. Updating R packages {#update-packages}

Open RStudio. In the R console, type the following and press Enter:

```{r eval=FALSE}
options(pkgType = "binary")
update.packages(ask = FALSE)
```





# History

- Last updated by UN in May 2024 for R version 4.4.0
- Last updated by UN in September 2023 for R version 4.3.1
- Last updated by UN in November 2022 for R version 4.2.2
- Last updated by UN in July 2022 for R version 4.2.1

Department of Psychology  
The University of Edinburgh
