---
title: "Updating R & Packages"
author: "uoepsy stats team"
output:
  html_document:
    theme: flatly
---

If your R version or packages are old, you might get errors or find that the code required to run some examples is different from the one provided.

Recall that R is the underlying programming language, while RStudio is just a wrapper around R providing a nicer user interface (text editor, file explorer and graphical window). Updating RStudio (the interface) will not update the R programming language which runs your code.  

You can find a video explaining how to update R, as well as step-by-step instructions below. 

<iframe id="kaltura_player" src="https://cdnapisec.kaltura.com/p/2010292/sp/201029200/embedIframeJs/uiconf_id/32599141/partner_id/2010292?iframeembed=true&playerId=kaltura_player&entry_id=1_8fywjo2v&flashvars[streamerType]=auto&amp;flashvars[localizationCode]=en&amp;flashvars[leadWithHTML5]=true&amp;flashvars[sideBarContainer.plugin]=true&amp;flashvars[sideBarContainer.position]=left&amp;flashvars[sideBarContainer.clickToClose]=true&amp;flashvars[chapters.plugin]=true&amp;flashvars[chapters.layout]=vertical&amp;flashvars[chapters.thumbnailRotator]=false&amp;flashvars[streamSelector.plugin]=true&amp;flashvars[EmbedPlayer.SpinnerTarget]=videoHolder&amp;flashvars[dualScreen.plugin]=true&amp;flashvars[Kaltura.addCrossoriginToIframe]=true&amp;&wid=1_d8xv05kr" width="304" height="231" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" frameborder="0" title="Kaltura Player"></iframe>


# 1. Check the version of R that you have

Check that your R version is up-to-date by typing into the R console:

```
version
```

After pressing Enter, you should see  something like this:

```
platform       <this will be different for each of you>     
arch           <this will be different for each of you>                  
os             <this will be different for each of you>     
system         <this will be different for each of you>        
status                                     
major          4                           
minor          0.2                         
[and more lines which you can ignore...]
```

Please check here: https://www.r-project.org/ to see what the most recent version is. If you have a different number for "major" than the most recent, it may well be time for an update.  

# 2. Update R

### Windows

Type the following into R pressing Enter to run the code:

```{r eval=FALSE}
install.packages("installr")
installr::updateR()
```


### Apple Mac

- Install XQuartz: [https://xquartz.macosforge.org/](https://xquartz.macosforge.org/)
- Go into Finder -> Applications -> Move R to trash (R, **not RStudio**)
- Download the most recent version of R for MacOS from [https://cran.r-project.org/bin/macosx/](https://cran.r-project.org/bin/macosx/) run the installer


# 3. Update the R packages

In R, type the following and press Enter:

```{r eval=FALSE}
options(pkgType = "binary")
update.packages(ask = FALSE)
```

